### 2019-10-31 manuscript/manuscript.rmd
keep updating

Over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested. In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occured early in animal evolution, espeically for complex structures such as cell types, nervous systems, and developmental patterns. However, a consensus has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have done here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

![](figures/Figure_overview.png)
**Fig XXOverview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) A ctenophore. (D) A sponge.

@@

### Gene sampling
Gene sampling has no doubt

### Outgroup taxon sampling

XXX

@@

Choanimalia, Holozoa, Opisthokonta

### Ingroup taxon sampling

Sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

ggplot(data=taxon_rectangles) +
	scale_x_continuous(name="x") +
	scale_y_continuous(name="y") +
	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
	facet_wrap( ~ manuscript_matrix )

@@

XXX

@@

ggplot( matrix_summary ) +
	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
	geom_abline(slope=1,intercept=0) +
	coord_fixed()

@@

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we reanalyze the primary matrices from each study under consistent conditions with iqtree. We selected this tool because it has greater model flexibility than other tools and is very fast.

We first tested a variety of models for each matrix, and inferred support under the selected model. We then analyzed every matrix under a panel of standard models, including XXX.

@@

## New analyses of new matrices

Based on the variation across analyses, we constructed new matrices with altered taxon and gene sampling to test specific hypotheses about differences in support.

@@

The count for a partition pair can be much arger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.

There are `r length(unique(partition_network_summary$matrix))` matrices. A gene that is perfectly sampled would form a cluster with this size. Very few clusters, though, are this size. This suggests that intersection of genes between matrices is low

>>>>

Phylogenomic analyses with genome-scale data have the potential to revolutionize our understanding of the tree of life. However, over the past decade there has been considerable debate about the position of the root of the animal phylogeny, with Ctenophora-sister (comb jelly) and Porifera-sister (sponges) (Fig XXOverview) emerging as the two primary hypotheses. Historically, there was little debate about the root of the animal tree of life and Porifera-sister was widely accepted though rarely tested since . In contrast to the lack of debate about the position of Porifera, there has long been uncertainty about the relationship of Ctenophora to other animals [@Wallberg:2004ws]. An understanding of the first diverging animal lineage is critical for scientists to reconstruct events that occurred  early in animal evolution, especially for complex structures such as cell types, nervous systems, and developmental patterns. However, a consensus has not been reached despite multiple efforts.

The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores are our most distant animal relative. Since then many more studies have been published, some supporting Ctenophora-sister, some Porifera-sister, and some neither. As it has become clear that this is a very difficult phylogenetic challenge, and the problem has become better characterized, it has become an interesting test-case to phylogenetic biologists beyond those concerned with this particular biological problem. Work has been hindered, though, because it has been difficult to directly compare results across studies and synthesize findings to understand the broader patterns of support. Here we synthesize data and results from all previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister, and reanalyze these data using standardized methods, and perform new analyses to characterize differences between studies. We hope that this effort provides an integrative overview of the challenge and provides direction for future studies. We also hope that the work we have conducted here, including consolidating all the datasets in one place with consistent formats and species names, will enhance the technical value of this interesting question to methods-focused investigators that look to develop methods to address difficult phylogenetic problems.

![](figures/Figure_overview.png)
**Fig XXOverview.** (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) A ctenophore. (D) A sponge. (E)

@@

### Gene sampling

Whole-genome and transcriptomic sequencing is leading to exciting new data matrices for phylogenomic analyses, potentially including hundreds or thousands of genes. Although the number of sampled genes might be different across studies, it is reasonable to expect that the sampled genes should be ultraconserved across animal genomes and most data matrices should share a core set of orthologs. Surprisingly, gene correspondance across different data matrices has been rarely assessed since most studies relied on their own bioinformatic pipeline for orthology determination.

### Outgroup and ingroup taxon sampling

Outgroup selection (the taxa used to root the ingroup animals) can potentially affect phylogenetic inference of the root position of animal phylogeny. Several studies have progressively removed more distantly related outgroup taxa to test the effect of outgroup selection to ingroup topology. Three progressively more inclusive clades have often been investigated, including Choanomalia (Metazoa plus most closely related Choanoflagellata), Holoza (Choanimalia plus more distantly related holozoans) and Opisthokonta (Holozoa + Fungi). It has been suggested the inclusion of more distant related (e.g. Opisthokonta) outgroup taxa tend to recover Ctenophora-sister using both site-homogeneous and site-heterogeneous models, whereas high support of Porifera-sister often recovered when site-heterogeneous models were used. In contrast of outgroup selection, sensitivity to ingroup sampling has received less attention than sensitivity to outgroup sampling. This may be because results have tended to be more sensitive to outgroup sampling.

@@

### Data-recoding

Compositional heterogeneity (lineage-specific differences in amino acid frequencies) was long thought has a major effect in resolving the animal tree of life, especially for the position of animal root. Feuda et al. 2017 has proposed to recode the full set of twenty amino acids into six groups of amino aicds that have more frequent evolutionary changes within groups than between groups based on different substitution models. By discarding information on which specific amino acid is found at each site in each species and instead focusing on which groups those amino acids belong to, they sought to reduce the impact of differences between species in which particular amino acids within those groups are most frequent. By reanalyzing Chang et al. 2015 and Whelan et al. 2015, they found that Porifera sister was favored under all recoding strategies using GTR-CAT models. However, the performance of the recoding methods has never been tested empirically against non-recoding method in the root of the animal phylogeny, and a recent simulation study has suggested that non-recoding approaches significantly outperformed 6-state recoding strategies.

@@

filtered_taxon_rectangles <- taxon_rectangles[!grepl("only", taxon_rectangles$matrix),]

ggplot(data=filtered_taxon_rectangles) +
	scale_x_continuous(name="x") +
	scale_y_continuous(name="y") +
	geom_rect( mapping=aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax, fill=clade )) +
	# scale_fill_manual(guide = guide_legend(reverse=TRUE))
	# scale_fill_manual(values = clade_colors, guide = guide_legend(reverse=TRUE))
	scale_fill_discrete(guide = guide_legend(reverse=TRUE)) +
	facet_wrap( ~ matrix )

@@

@@

p1 = ggplot( matrix_summary ) +
	geom_point( aes(x=n_partitions, y=n_busco_partitions, col=manuscript) ) +
	geom_abline(slope=1,intercept=0) +
	coord_fixed()

matrix_summary$matrix = factor( matrix_summary$matrix, levels=c("Dunn2008", "Philippe2009", "Hejnol2009", "Nosenko2013_nonribo_9187","Nosenko2013_ribo_11057","Nosenko2013_ribo_14615","Ryan2013_est", "Moroz2014_3d", "Chang2015", "Borowiec2015_Best108" , "Borowiec2015_Total1080", "Whelan2015_D1", "Whelan2015_D10", "Whelan2015_D20", "NA", "Whelan2017_full", "Whelan2017_strict") )

p2 = ggplot( matrix_summary ) + geom_point( aes(x=matrix, y=n_busco_partitions/n_partitions*100, col=manuscript) ) + 	theme_classic() + theme(axis.text.x = element_text(angle=45, hjust=1))
```

@@

One of the challenges of interpreting support for the placement of the animal root across studies is that different programs, software versions, and settings have been used across studies, and phylogenetic analysis decisions have been approached in very different ways. Here we reanalyze the primary matrices from each study under consistent conditions with IQtree under a panel of evolutionary models. We selected this tool because it has greater model flexibility, computational time and accuracy than other tools in maximum likelihood framework.

We first tested a variety of models for each matrix, and inferred support under the best-fit model from IQtree. We then analyzed every matrix under a panel of standard models, including WAG+G, GTR+G, Poisson+G.

@@

@@

The count for a partition pair can be much larger than the number of genes in the matrix, which suggests that the count is the number of hsps rather than the number of sequences with hits.

There are `r length(unique(partition_network_summary$matrix))` matrices. A gene that is perfectly sampled would form a cluster with this size. To our surprise, very few clusters, though, are this size. This suggests that intersection of genes between matrices is extreme low. This finding suggested genes used for exploring the position of the root of the animal phylogeny
